Improving reproducibility of scientific software using Nix/NixOS: A case study on the preCICE ecosystem

### 2024-05-14 paper/paper.tex
Add funding acknowledgment
Rewrite discussion

After detailing the challenges of building and packaging each solver and their respective adapter, we want to discuss some common points between these scientific software packages.
preCICE, as a multi-component research software, covers many different simulation codes and therefore provides a good overview over different kinds of packages, making this case study representative for multi-component research software.\\

Generally speaking, there are major gaps regarding the ability to successfully package the software between each different software solution that can be traced back to the build system.
Custom build systems require an disproportional amount of additional work to make them runnable on any system.
OpenFOAM and Allwmake are prime examples of these issues because now package managers have to understand a whole new build system for exactly one package in order to successfully package it and make it easily available to users.
It is also worth mentioning that writing your own custom build system, requires additional time to maintain this build system, that could have been spent working on the software itself.
Packages that use common, industry proven package managers, such as CMake and autotools make it easier for developers of the software to build their software as well as package managers to package it.
These build tools, generally also provide interfaces for dependency management that can be used to tell users and package maintainers what dependencies are needed, and it is usually better than only documenting required dependencies, because documentation needs to be kept in sync with the underlying code.
They also generally already provide proven interfaces for configuring packages with additional features, and these also might be self documenting.

The authors of the xSDK\footnote{\url{https://xsdk.info/}} came to a very similar conclusion.
Their goal is to ``identify, adapt, and adopt best practices in software engineering'' and ``to provide the foundation of this extensible scientific software ecosystem''~\footnote{\url{https://xsdk.info/}}.
To achieve this they provide a list of package policies~\cite{xSDK2023}, that for example specifies as mandatory policy that packages ``must support portable installation through Spack''~\cite{xSDK2023} and that all packages ``should have a build system that is appropriate for the language''~\cite{xSDK2023}, that includes CMake and autoconf as examples.
If a package supports all these policies, they can be added to the growing list of packages in the xSDK, that currently among other things include PETSc, deal.ii and preCICE.
These packages are already part of nixpkgs, or it is easy to provide a package for them because they support portable installation, meaning installation into custom prefixes such as nix store paths, and common build systems such as CMake and autoconf.

Another finding is, that is also addressed in xSDK through portable installations, is the requirement to install software to be installed into \texttt{\$HOME} or requires files to be located at a specific location on your file system.
Most of the time, it is possible to work around this, but there are already available solutions for dependency management.
Both, the CalculiX-adapter and SU2-adapter do require the original source code to be present as they add new features on top or patch the source code.
The current solution of the user cloning the code at the specific location works, but it might be better to look into submodules because that also allows the developers to pin the version more easily rather than documenting it, as that could more easily lead to user error, when they try to package and install the adapters.

Additionally, hard-coding libraries to \texttt{/usr/lib} and \texttt{/usr/include} might work on one distribution, but might make it hard or even impossible to install a package on a different distribution or even a later version of the same distribution.
Solutions such as \texttt{pkg-config} that can be configured, so it finds libraries outside the common paths, should always be preferred and work out of the box with the common paths such as \texttt{/usr/lib}.

With Nix' dependency management, we can easily copy the package definitions we just wrote onto different systems, e.g., entirely different Linux distributions, and execute the build of the closures there.
Building the same closures yields the same outputs, as already described in section \ref{sec:nix}.

@@

>>>>

After detailing the challenges of building and packaging all components of the preCICE ecosystem and running the tutorials as testcases, we want to discuss common points between these scientific software packages.
The required effort to package the individual components varies drastically. Major problems can be traced back to the build system.
Custom build systems require an disproportional amount of additional work to make them runnable on any system.
OpenFOAM and Allwmake are prime examples of this issue. Package managers have to understand a whole new build system for a single package -- a fact that is often ignored when considering the trade-off between maintaining a custom build system versus pivoting to an industry standard, such as CMake or autotools.
Such standard build tools generally provide interfaces for dependency management and optional features. This is often better than only documenting these, since documentation needs to be kept in sync with the underlying code.

The authors of xSDK\footnote{\url{https://xsdk.info/}} come to a similar conclusion.
They provide a list of package policies~\cite{xSDK2023}, which, for example, specify that packages must have an appropriate build system, which includes CMake and autoconf as examples.
If a package supports all policies, they can be added to the growing list of packages in xSDK.
It is no surprise that xSDK packages (e.g., PETSc, deal.II, or preCICE) are also easy to package with Nix.

The xSDK policies also requires portable installations. Several components in the preCICE ecosystem require installation into \texttt{\$HOME} or any required files to be located at specific locations. Both, the CalculiX-preCICE adapter and the SU2-preCICE adapter require the original source code to be present as they add new features on top or patch the source code.
The current solution of cloning the code at the specific location works, but it might be better to look into Git submodules. This would moreover allow pinning the version of the original code compared to only documenting it.

Finally, hard-coding libraries to \texttt{/usr/lib} and \texttt{/usr/include} might work on one system, but might make it hard or even impossible to install a package on a different system.
Configurable solutions, such as \texttt{pkg-config}, should always be preferred and work out of the box with common paths such as \texttt{/usr/lib}.

@@

\acknowledge{
We thankfully acknowledge the funding by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) under Germany's Excellence Strategy EXC 2075 -- 390740016 and the support by the Stuttgart Center for Simulation Science (SimTech).
}
